\documentclass{article}
\usepackage{graphicx}

\title{AutoCTM Software Specification}
\author{Oscar Ashburn}
\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
This document specifies the functional and technical requirements for AutoCTM 
(Automated Certificate Transparency Monitor), a tool written in Go for 
monitoring Certificate Transparency (CT) logs. It defines the system's 
expected behaviour, interfaces, and constraints, serving as the authoritative 
reference for the design and implementation of AutoCTM.
\end{abstract}
\newpage

\section{Introduction}

\subsection{What is Certificate Transparency}
Certificate Transparency (CT) is a system designed to make TLS/SSL certificate 
issuance public, auditable, and verifiable. It helps prevent mis-issued or 
rogue certificates and increases trust in HTTPS.

\subsection{The Problem}
Certificate Authorities (CAs) issue TLS/SSL certificates for domains. 
Sometimes, a CA can mistakenly issue a certificate to the wrong party, or a 
malicious actor could obtain one. Without transparency, domain owners might 
not know about these certificates until it is too late.

\begin{itemize}
    \item CAs can mistakenly or maliciously issue certificates for domains 
    they should not.
    \item Domain owners may have no visibility into certificates issued in 
    their name.
    \item The window of unawareness is the dangerous part. An attacker could 
    use a fraudulent certificate before anyone notices.
\end{itemize}

\subsection{How CT Works}
Certificate Transparency solves the problem of rogue certificate issuance by 
requiring every CA to submit certificates to public, append-only logs before 
browsers will trust them. This means anyone can monitor these logs to spot 
unauthorised certificates for their domain, and browsers will outright reject 
any certificate that lacks proof of being logged.

\section{Design Description}
AutoCTM is written in Go and is composed of three decoupled processes: a 
persistent broker daemon (\texttt{autoctm-broker}), one or more monitor 
instances (\texttt{autoctm-instance}), and a lightweight operator CLI 
(\texttt{autoctm}). The system must support the following requirements:

\begin{itemize}
    \item \textbf{Broker Daemon.} The system shall run a persistent background 
    broker process that owns the control socket, manages the instance registry, 
    and routes commands between the CLI and monitor instances. The broker shall 
    start automatically on first CLI launch and remain running independently 
    of the CLI.

    \item \textbf{CLI Interface.} The system shall provide a command-line 
    interface that connects to the broker on launch, accepts operator commands, 
    and may be closed at any time without affecting running instances. On 
    relaunch the CLI shall reconnect to the existing broker and resume control.

    \item \textbf{Monitor Instances.} The system shall support spawning one or 
    more independent monitor instances as background processes. Each instance 
    shall monitor one or more RFC 9162-compliant CT logs, polling for new 
    entries, verifying STH signatures, and validating Merkle consistency proofs 
    between successive Signed Tree Heads.

    \item \textbf{Certificate Database.} The system shall persist all observed 
    certificate entries to a shared PostgreSQL database, recording at minimum 
    the leaf hash, log ID, timestamp, subject, SANs, issuer, and serial number 
    for each entry.

    \item \textbf{Certificate Revocation Checking.} The system shall check 
    observed certificates against current Certificate Revocation Lists (CRLs) 
    and flag any certificate whose serial number appears as revoked.

    \item \textbf{Domain Watchlist Alerting.} The system shall allow operators 
    to configure a shared watchlist of domains and shall raise an alert whenever 
    a certificate covering a watched domain is observed in any monitored log.
\end{itemize}

\subsection{System Features}

\subsubsection{Broker Daemon}
AutoCTM shall run a persistent broker daemon that acts as the backbone of 
the system. The broker shall:
\begin{itemize}
    \item Own and maintain a Unix domain socket at a well-known path, used 
    for all CLI-to-broker and broker-to-instance communication.
    \item Maintain an instance registry tracking each instance's ID, PID, 
    and current status.
    \item Persist the instance registry to the database so that it survives 
    a broker restart.
    \item On restart, reload the instance registry and attempt to reconnect 
    to any instances whose process is still active.
    \item Spawn new monitor instance processes on operator request.
    \item Route commands from the CLI to the appropriate instance.
    \item Start automatically when the CLI is launched and no broker is found.
\end{itemize}

\subsubsection{CLI Interface}
AutoCTM shall provide a command-line interface for operator control. The CLI 
shall:
\begin{itemize}
    \item On launch, connect to the broker socket. If no broker is found, 
    automatically spawn one before presenting the command prompt.
    \item Operate in one of two contexts: \textit{global context} (no instance 
    selected) and \textit{instance context} (an active instance is selected 
    via \texttt{/set-instance}).
    \item Support the following commands in global context:
    \begin{itemize}
        \item \texttt{/start} --- Spawn a new monitor instance.
        \item \texttt{/get-instances} --- List all instances and their status.
        \item \texttt{/set-instance <instanceID>} --- Set the active instance. 
        Subsequent commands are routed to this instance.
        \item \texttt{/add <domain>} --- Add a domain to the shared watchlist.
        \item \texttt{/remove <domain>} --- Remove a domain from the shared 
        watchlist.
        \item \texttt{/shutdown} --- Gracefully shut down all instances and 
        the broker.
    \end{itemize}
    \item Support the following commands in instance context:
    \begin{itemize}
        \item \texttt{/add-log <url>} --- Add a CT log to the instance.
        \item \texttt{/remove-log <url>} --- Remove a CT log from the instance.
        \item \texttt{/pause} --- Pause the instance's polling loop.
        \item \texttt{/resume} --- Resume a paused instance.
        \item \texttt{/status} --- Display the instance's current status.
        \item \texttt{/shutdown} --- Shut down the active instance only.
        \item \texttt{/clear-instance} --- Return to global context.
    \end{itemize}
    \item May be closed at any time. Running instances and the broker shall 
    be unaffected. On relaunch the CLI shall reconnect to the existing broker.
\end{itemize}

\subsubsection{CT Log Monitoring}
Each AutoCTM monitor instance shall monitor one or more RFC 9162-compliant 
CT logs. The instance shall:
\begin{itemize}
    \item Poll each configured log for a new STH on a configurable interval.
    \item Verify the STH signature before processing any entries.
    \item Fetch all new entries since the last known STH.
    \item Verify Merkle consistency proofs between successive STHs.
    \item Support monitoring of multiple logs concurrently via independent 
    goroutines.
    \item Detect and report log misbehaviour including MMD violations, 
    unavailable entries, and consistency proof failures.
\end{itemize}

\subsubsection{Certificate Storage}
AutoCTM shall persist all observed certificate entries to a shared PostgreSQL 
database. The system shall:
\begin{itemize}
    \item Store the leaf hash, log ID, timestamp, subject, SANs, issuer, 
    serial number, and expiry for each observed certificate.
    \item Prevent duplicate entries for the same leaf hash.
    \item Support querying stored certificates by domain, issuer, and 
    date range.
\end{itemize}

\subsubsection{Certificate Revocation Checking}
AutoCTM shall check observed certificates against Certificate Revocation 
Lists (CRLs). The system shall:
\begin{itemize}
    \item Fetch and parse CRLs for each issuer observed in monitored logs.
    \item Refresh CRLs on a configurable interval.
    \item Flag and record any certificate whose serial number appears in 
    a current CRL.
\end{itemize}

\subsubsection{Domain Watchlist Alerting}
AutoCTM shall allow operators to configure a shared watchlist of domains to 
monitor for unexpected certificate issuance. The system shall:
\begin{itemize}
    \item Match observed certificates against the watchlist by exact domain 
    and wildcard SAN.
    \item Raise an alert for any certificate covering a watched domain.
    \item Allow watchlist entries to be added and removed via the CLI without 
    restarting any instance or the broker.
\end{itemize}

\subsubsection{Log Integrity Auditing}
AutoCTM shall detect and report misbehaviour in monitored logs as defined 
in RFC 9162. The system shall:
\begin{itemize}
    \item Detect and report STHs that exceed the log's declared Maximum 
    Merge Delay (MMD).
    \item Report entries that remain unavailable for an extended period 
    after appearing in an STH.
    \item Record and report any consistency proof verification failure.
    \item Pause polling of a log on consistency proof failure pending 
    operator review.
\end{itemize}

\subsubsection{Alert Notification}
AutoCTM shall deliver alerts to operators via configurable notification 
channels. The system shall:
\begin{itemize}
    \item Support at least one outbound notification channel 
    (e.g. email or webhook).
    \item Include in each alert the affected certificate details, the 
    triggering log, and the reason for the alert.
    \item Avoid duplicate alerts for the same certificate and alert type.
\end{itemize}

\subsection{Non-Functional Requirements}

\end{document}