\documentclass{article}
\usepackage{geometry}
\geometry{margin=1in}

\title{AutoCTM Software Design Specification}
\author{}
\date{}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section{Introduction}

\subsection{Purpose}
This document describes the software design of AutoCTM, a Certificate
Transparency monitor written in Go. It is intended to guide implementation
and serve as a reference for the development team.

\subsection{Scope}
AutoCTM allows operators to monitor one or more RFC 9162-compliant CT logs
for certificate issuance against a configurable domain watchlist. The system
consists of three decoupled components: a CLI, a persistent broker daemon,
and one or more independent monitor instances. The CLI may be closed at any
time without affecting running instances. On relaunch the CLI reconnects to
the broker and resumes control.

\subsection{Known Limitations}
\subsubsection{Broker as Single Point of Failure}
The broker is a single point of failure for the control plane. If the broker
dies:
\begin{itemize}
    \item The CLI cannot connect or issue commands.
    \item Instance status becomes unknown until the broker is restarted.
\end{itemize}
This is an accepted limitation for a small internal tool. Instances are
designed to continue monitoring and writing to the database independently of
the broker. Operators can restart the broker at any time to resume control
without affecting running instances.

\subsection{Definitions}
\begin{itemize}
    \item \textbf{CLI} --- The operator-facing command-line interface. A
    short-lived process that connects to the broker, issues commands, and
    may be closed freely.
    \item \textbf{Broker} --- A persistent background daemon that owns the
    control socket, manages the instance registry, and routes commands from
    the CLI to the appropriate instance. The broker is control plane only ---
    it does not participate in certificate monitoring.
    \item \textbf{Instance} --- A long-running background process responsible
    for monitoring one or more CT logs, checking certificates, and raising
    alerts. Instances load their configuration directly from the database and
    operate independently of the broker at runtime.
    \item \textbf{Active Instance} --- The instance currently selected in the
    CLI via \texttt{/set-instance}. Commands issued in instance context are
    routed to this instance.
    \item \textbf{STH} --- Signed Tree Head, as defined in RFC 9162.
    \item \textbf{SCT} --- Signed Certificate Timestamp, as defined in RFC 9162.
    \item \textbf{MMD} --- Maximum Merge Delay, as defined in RFC 9162.
    \item \textbf{Watchlist} --- The set of domains configured by the operator
    for alert monitoring.
\end{itemize}

\subsection{References}
\begin{itemize}
    \item RFC 9162 --- Certificate Transparency Version 2.0
    \item AutoCTM Software Requirements Specification
\end{itemize}

\section{System Overview}

AutoCTM is composed of three decoupled processes:

\begin{itemize}
    \item \textbf{CLI} (\texttt{autoctm}) --- The operator entry point. On
    launch it checks for a running broker. If none is found it spawns one
    automatically before presenting the command prompt. The CLI may be closed
    at any time; running instances are unaffected.
    \item \textbf{Broker} (\texttt{autoctm-broker}) --- A persistent daemon
    that starts automatically on first CLI launch. It owns a local control
    socket at a well-known path, maintains the instance registry, and routes
    commands between the CLI and instances. It remains alive independently of
    both the CLI and any instance. The broker is control plane only --- loss
    of the broker does not affect instance monitoring.
    \item \textbf{Monitor Instance} (\texttt{autoctm-instance}) --- A
    long-running background process spawned by the broker on \texttt{/start}.
    Each instance loads its configuration directly from the database,
    independently polls its configured CT logs, processes entries, performs
    watchlist and CRL checks, and raises alerts. Multiple instances may run
    concurrently. Instances continue to operate if the broker becomes
    unavailable.
\end{itemize}

\subsection{Separation of Responsibilities}
To avoid ambiguity over configuration ownership, the system follows a clear
separation of responsibilities:
\begin{itemize}
    \item \textbf{Database} --- The single source of truth for all
    configuration and state, including instance config, log lists, watchlist,
    STH state, and alerts.
    \item \textbf{Broker} --- Orchestration only. Spawns instances, maintains
    the instance registry, and routes control commands. Does not own
    configuration.
    \item \textbf{Instance} --- Worker. Loads all configuration directly from
    the database on startup and re-reads it when notified of a change by the
    broker.
\end{itemize}

\section{Architecture}

\subsection{Component Diagram}
% Insert component diagram here

\subsection{Process Lifecycle}

\subsubsection{Broker Startup}
\begin{enumerate}
    \item CLI launches and checks for a broker socket at the well-known path.
    \item If no socket is found, the CLI forks and execs \texttt{autoctm-broker}
    as a detached background process, then waits for the socket to appear.
    \item The broker creates the socket, loads the instance registry from the
    database, and begins listening for CLI and instance connections.
\end{enumerate}

\subsubsection{CLI Lifecycle}
\begin{enumerate}
    \item On launch, the CLI connects to the broker socket.
    \item The CLI enters a REPL, accepting operator commands and forwarding
    them to the broker.
    \item The CLI maintains an \textit{active instance} context, initially
    unset (global context).
    \item The CLI may be closed at any time with no effect on the broker or
    any running instance. On relaunch it reconnects to the existing broker.
\end{enumerate}

\subsubsection{Instance Lifecycle}
\begin{enumerate}
    \item The broker spawns a new instance process on receipt of
    \texttt{/start}, passing only the instance ID.
    \item The instance connects to the database and loads its full
    configuration directly from the \texttt{instances} table.
    \item The instance attempts to register with the broker via the control
    socket. If the broker is unavailable, the instance shall continue startup
    and retry registration periodically.
    \item The instance spawns one goroutine per configured log and enters
    its polling loop.
    \item On \texttt{/shutdown} (targeted or global), the broker signals the
    instance to drain its current work and exit cleanly.
    \item On exit, the instance updates its status to \texttt{stopped} in
    the database.
\end{enumerate}

\subsubsection{Broker Restart Behaviour}
If the broker is restarted it shall:
\begin{enumerate}
    \item Recreate the control socket at the well-known path.
    \item Reload the instance registry from the database.
    \item Wait for instances to reconnect and re-register. The broker shall
    not attempt to reconnect to instances --- instances are responsible for
    detecting the lost connection and re-registering with the restarted
    broker.
    \item Mark any instance that does not re-register within a configurable
    timeout as \texttt{stopped}.
\end{enumerate}

\subsection{Inter-Process Communication}
All IPC uses a local control socket owned by the broker at a well-known
path. Messages are encoded as newline-delimited JSON. There are two
connection types:

\begin{itemize}
    \item \textbf{CLI $\rightarrow$ Broker} --- The CLI holds a persistent
    connection to the broker for the duration of its session.
    \item \textbf{Instance $\rightarrow$ Broker} --- Each instance initiates
    and maintains a connection to the broker for registration and command
    receipt. If the broker connection is lost, the instance shall keep
    running and retry the connection periodically until the broker is
    available again.
\end{itemize}

\section{Component Design}

\subsection{CLI}

\subsubsection{Command Contexts}
The CLI operates in one of two contexts at any time:

\paragraph{Global Context}
No active instance is selected. Commands affect shared state or the broker:
\begin{itemize}
    \item \texttt{/start} --- Instruct the broker to spawn a new monitor
    instance. Prints the assigned instance ID.
    \item \texttt{/get-instances} --- List all registered instances, their
    IDs, and current status.
    \item \texttt{/set-instance <instanceID>} --- Set the active instance.
    The CLI prompt updates to reflect the selected instance. Subsequent
    commands are routed to that instance.
    \item \texttt{/add <domain>} --- Add a domain to the shared watchlist.
    \item \texttt{/remove <domain>} --- Remove a domain from the shared
    watchlist.
    \item \texttt{/shutdown} --- Gracefully shut down all instances and the
    broker. The CLI then exits.
\end{itemize}

\paragraph{Instance Context}
An active instance is selected via \texttt{/set-instance}. Commands are
routed to that instance:
\begin{itemize}
    \item \texttt{/add-log <url>} --- Add a CT log to the instance's
    configuration. The broker writes the new log to the database and
    notifies the instance to reload its configuration.
    \item \texttt{/remove-log <url>} --- Remove a CT log from the instance's
    configuration.
    \item \texttt{/pause} --- Pause the instance's polling loop.
    \item \texttt{/resume} --- Resume a paused instance.
    \item \texttt{/status} --- Print the instance's current status, including
    configured logs, last known STH per log, and alert counts.
    \item \texttt{/shutdown} --- Shut down the active instance only.
    \item \texttt{/clear-instance} --- Return to global context.
\end{itemize}

\subsubsection{Auto-Spawn Behaviour}
If the CLI cannot connect to the broker socket on launch it shall
automatically spawn the broker as a detached background process before
presenting the prompt. This requires no manual daemon management from the
operator.

\subsection{Broker}

\subsubsection{Responsibilities}
The broker is responsible for orchestration only. It shall:
\begin{itemize}
    \item Own and maintain the local control socket.
    \item Maintain an in-memory instance registry of connected instances,
    backed by the database as the source of truth.
    \item Route CLI commands to the correct instance.
    \item Spawn new instance processes on \texttt{/start}, writing the new
    instance record to the database before spawning.
    \item Accept re-registration from instances after a broker restart.
    \item Write instance status changes to the database.
\end{itemize}

The broker shall not hold instance configuration in memory. All configuration
reads and writes go through the database.

\subsection{Monitor Instance}

\subsubsection{Startup Sequence}
\begin{enumerate}
    \item Receive instance ID from the broker at spawn time.
    \item Connect to PostgreSQL and load full configuration from the
    \texttt{instances} table using the instance ID.
    \item Attempt to register with the broker. If unavailable, continue
    and retry in the background.
    \item Spawn one goroutine per configured log.
    \item Enter polling loop.
\end{enumerate}

\subsubsection{Broker Connection Resilience}
Instances shall treat the broker connection as optional at runtime:
\begin{itemize}
    \item If the broker connection is lost, the instance shall continue
    its polling loop uninterrupted.
    \item The instance shall periodically retry connecting and registering
    with the broker using an exponential backoff.
    \item On successful reconnection, the instance shall re-register and
    resume accepting commands.
\end{itemize}

\subsubsection{Polling Loop}
Each log goroutine shall independently perform the following steps:
\begin{enumerate}
    \item Fetch the current STH via \texttt{GET /ct/v2/get-sth}.
    \item Verify the STH signature using the log's public key.
    \item If the STH is unchanged from the stored STH, wait for the
    configured poll interval and retry.
    \item Fetch new entries since the last known STH via
    \texttt{GET /ct/v2/get-entries}.
    \item For each entry, perform watchlist and CRL checks.
    \item Verify the Merkle consistency proof between the previous and new
    STH via \texttt{GET /ct/v2/get-sth-consistency}.
    \item Persist new entries and update the stored STH in the database.
    \item Dispatch any pending alerts via email.
    \item Wait for the configured poll interval and return to step 1.
\end{enumerate}

\subsubsection{Watchlist Checking}
Each certificate's Subject CN and SAN fields are matched against all domains
in the watchlist. Matching shall support:
\begin{itemize}
    \item Exact domain match (e.g. \texttt{example.com}).
    \item Wildcard SAN match (e.g. \texttt{*.example.com}).
\end{itemize}
A match shall produce an alert record of type \texttt{watchlist\_match}.

\subsubsection{CRL Checking}
The instance shall maintain a local in-memory cache of CRLs keyed by issuer
DN. CRLs shall be refreshed on a configurable interval. Each observed
certificate serial number shall be checked against the cached CRL for its
issuer. A hit shall produce an alert record of type \texttt{revoked}.

\section{Data Design}
\label{sec:data}

\subsection{Database}
AutoCTM uses a local PostgreSQL instance as its storage backend. The database
is the single source of truth for all configuration and state. It is shared
across the broker and all monitor instances.

\subsection{Schema}

\subsubsection{certificates}
Stores all observed certificate entries.
\begin{itemize}
    \item \texttt{id} --- Primary key (UUID).
    \item \texttt{leaf\_hash} --- SHA-256 leaf hash (unique).
    \item \texttt{log\_id} --- OID of the source log.
    \item \texttt{observed\_at} --- Timestamp of observation.
    \item \texttt{not\_before}, \texttt{not\_after} --- Certificate validity
    period.
    \item \texttt{subject} --- Certificate subject CN.
    \item \texttt{sans} --- Array of Subject Alternative Names.
    \item \texttt{issuer} --- Issuer DN.
    \item \texttt{serial} --- Certificate serial number.
    \item \texttt{revoked} --- Boolean, updated by CRL checks.
\end{itemize}

\subsubsection{watchlist}
\begin{itemize}
    \item \texttt{id} --- Primary key (UUID).
    \item \texttt{domain} --- Domain pattern (exact or wildcard).
    \item \texttt{added\_at} --- Timestamp of addition.
\end{itemize}

\subsubsection{instances}
Stores instance identity, status, and configuration. This is the source of
truth for instance configuration --- instances load directly from this table
on startup.
\begin{itemize}
    \item \texttt{id} --- Instance ID (UUID).
    \item \texttt{pid} --- OS process ID.
    \item \texttt{status} --- Current status (\texttt{running},
    \texttt{paused}, \texttt{stopped}).
    \item \texttt{started\_at} --- Timestamp.
    \item \texttt{config} --- JSONB blob of instance configuration (log
    URLs, public keys, poll interval). Written by the broker, read by
    the instance.
\end{itemize}

\subsubsection{sth\_state}
Tracks the last known STH per log per instance.
\begin{itemize}
    \item \texttt{instance\_id} --- Foreign key to \texttt{instances}.
    \item \texttt{log\_id} --- OID of the log.
    \item \texttt{tree\_size} --- Tree size of the last known STH.
    \item \texttt{root\_hash} --- Root hash of the last known STH.
    \item \texttt{timestamp} --- STH timestamp.
\end{itemize}

\subsubsection{alerts}
\begin{itemize}
    \item \texttt{id} --- Primary key (UUID).
    \item \texttt{certificate\_id} --- Foreign key to \texttt{certificates}.
    \item \texttt{instance\_id} --- Foreign key to \texttt{instances}.
    \item \texttt{reason} --- Alert reason (\texttt{watchlist\_match},
    \texttt{revoked}, \texttt{log\_misbehaviour}).
    \item \texttt{detail} --- JSONB blob of alert-specific detail.
    \item \texttt{raised\_at} --- Timestamp.
    \item \texttt{dispatched} --- Boolean, whether the notification was sent.
\end{itemize}

\section{Error Handling}

\subsection{Log Unavailability}
If a CT log is unreachable the instance shall log the failure and back off
using exponential retry. If the log remains unavailable beyond the MMD the
instance shall raise a \texttt{log\_misbehaviour} alert.

\subsection{Consistency Proof Failure}
A consistency proof verification failure shall be treated as log
misbehaviour. The instance shall record the failure, raise an alert, and
pause polling of the affected log pending operator review via \texttt{/resume}.

\subsection{Database Connectivity}
If the database becomes unreachable the instance shall pause processing and
retry on a backoff schedule. On recovery the instance shall resume from the
last persisted STH state, re-fetching any entries missed during the outage.

\subsection{Broker Unavailability}
If the CLI cannot reach the broker it shall attempt to auto-spawn a new
broker instance. If the broker cannot be spawned the CLI shall print a
diagnostic error and exit.

If a running instance loses its connection to the broker it shall continue
monitoring uninterrupted and retry registration on an exponential backoff
until the broker is available again.

\subsection{Instance Crash}
If the instance crashes without a clean exit, the broker shall detect the lost
connection and mark the instance as \texttt{stopped} in the database.

If the broker detects that an instance connection has been lost unexpectedly,
it shall mark the instance as stopped in the database.

\end{document}